---
title: "Class17: Vaccination Rate Mini Project"
author: "Ashlyn Murphy"
format: gfm
---

# Background

Today's class is exploratory data analysis on statewide vaccination rates posted by CA.gov

The goal of this hands-on mini-project is to examine and compare the Covid-19 vaccination rates around San Diego.

We will start by downloading the most recently dated “Statewide COVID-19 Vaccines Administered by ZIP Code” CSV file from: https://data.ca.gov/dataset/covid-19-vaccine-progress-dashboard-data-by-zip-code

## Data Input


```{r}
vax <- read.csv("covid19vaccinesbyzipcode_test.csv")

head(vax)
```

```{r}
skimr::skim(vax)
```

> Q1. What column details the total number of people fully vaccinated?

```{r}
head(vax[10])
```
column 10 details people fully vaccinated

> Q2. What column details the Zip code tabulation area? 

```{r}
head(vax[2])
```
column 2 is the zip code tabulated area

> Q3. What is the earliest date in this dataset?

```{r}
min(vax[]$as_of_date)
```

the earliest date is 1/5/2021

> Q4. What is the latest date in this dataset?

```{r}

max(vax[]$as_of_date)
```

the last date is 3/07/2023

> Q5. How many numeric columns are in this dataset?

from the skimr package we can see there are 13 numeric columns

> Q6. Note that there are “missing values” in the dataset. How many NA values there in the persons_fully_vaccinated column?

from the skimr package we can see there are 16621 NA values in the persons_fully_vaccinated or use the following code

```{r}
total_vax <- sum(is.na(vax$persons_fully_vaccinated))
total_vax
```

> Q7. What percent of persons_fully_vaccinated values are missing (to 2 significant figures)?

```{r}
round((total_vax/nrow(vax))*100, 2)
```

# Working with dates

use the `lubridate` package to help you work with times and dates

```{r}
library(lubridate)
```

```{r}
today()
```

```{r}
#today() - vax$as_of_date[1]
```
 ^ doesnt work bc as_of_date column is actually in character strings which cant be used for math

```{r}
ymd("1999-4-14")
```

```{r}
today() - ymd("1999-4-14")
```

```{r}
vax$as_of_date <- ymd(vax$as_of_date)
```


```{r}
today() - vax$as_of_date[1]
```

days since the first vaccination: 792

> Q9. How many days have passed since the last update of the dataset?

```{r}
today() - vax$as_of_date[nrow(vax)]

```

its been 1 day since it was last updated

> Q10. How many unique dates are in the dataset (i.e. how many different dates are detailed)? 

```{r}
length(unique(vax$as_of_date))
```

or use a dplyr function

```{r}
library(dplyr)

n_distinct(vax$as_of_date)
```

# Working with Zip codes

Many packages exist to work with zip codes, we will use `zipcodeR`

```{r}
library(zipcodeR)
```

```{r}
geocode_zip('92037')
```

```{r}
geocode_zip('48044')
```

```{r}
zip_distance('92037','48044')
```
```{r}
zip <- reverse_zipcode(c('92037', "48044") )

zip
```


find the best ratio of median household income in San Diego

first find all san diego zip codes

> Q11. How many distinct zip codes are listed for San Diego County?

```{r}
sd.zip <- unique(vax$zip_code_tabulation_area[vax$county == "San Diego"])
```

```{r}
length(sd.zip)
```

107 distinct zips in SD county

now do a zip code look up for the data we want

```{r}
sd.eco <- reverse_zipcode(sd.zip)

sd.eco
```

and extract the values we want, which is the median_household_income and median_home_value


most expensive area: 

```{r}
sd.eco$median_household_income
```

```{r}
ord <- order(sd.eco$median_home_value, decreasing = T)

head(sd.eco[ord,])
```

```{r}
arrange(sd.eco, desc(median_home_value))
```

least expensive 

```{r}
arrange(sd.eco, median_home_value)
```

using dplyr...

```{r}
sd <- filter(vax, county == "San Diego")

nrow(sd)
```

```{r}
sd.10 <- filter(vax, county == "San Diego" &
                age5_plus_population > 10000)

nrow(sd.10)
```

> Q12. What San Diego County Zip code area has the largest 12 + Population in this dataset?

```{r}
ind <- which.max(sd$age12_plus_population)
sd[ind,]
```



> Q13. What is the overall average “Percent of Population Fully Vaccinated” value for all San Diego “County” as of “2023-03-07”?

```{r}
sd_latest<- filter(sd, as_of_date == "2023-03-07")

head(sd_latest)

```

```{r}
mean(sd_latest$percent_of_population_fully_vaccinated, na.rm = T)
```



> Q14. Using either ggplot or base R graphics make a summary figure that shows the distribution of Percent of Population Fully Vaccinated values as of “2023-03-07”?


```{r}
library(ggplot2)
```

```{r}
ggplot(sd_latest) +
  aes(sd_latest$percent_of_population_fully_vaccinated) +
  geom_histogram(bins = 20)
```

#UCSD/La Jolla Area

```{r}
ucsd <- filter(sd, zip_code_tabulation_area=="92037")
ucsd[1,]$age5_plus_population
```

> Q15. Using ggplot make a graph of the vaccination rate time course for the 92037 ZIP code area:

```{r}
ljplot <- ggplot(ucsd) +
  aes(ucsd$as_of_date, ucsd$percent_of_population_fully_vaccinated) +
  geom_point() +
  geom_line(group = 1) +
  ylim(c(0,1)) +
  labs(x = "Date", y = "Percent Vaccinated")

ljplot
```

> Q16. Calculate the mean “Percent of Population Fully Vaccinated” for ZIP code areas with a population as large as 92037 (La Jolla) as_of_date “2023-02-28”. Add this as a straight horizontal line to your plot from above with the geom_hline() function?

```{r}
vax.36 <- filter(vax, age5_plus_population > 36144 &
                as_of_date == "2023-03-07")

head(vax.36)
```

First figure out how many areas are at least as big as LJ & calc the percent

```{r}
avg.36 <- mean(vax.36$percent_of_population_fully_vaccinated, na.rm = T)

avg.36

```

```{r}
ljplot + geom_hline(yintercept = avg.36, col = "red", linetype = 2)
```



> Q17. What is the 6 number summary (Min, 1st Qu., Median, Mean, 3rd Qu., and Max) of the “Percent of Population Fully Vaccinated” values for ZIP code areas with a population as large as 92037 (La Jolla) as_of_date “2023-02-28”?

```{r}
summary(vax.36)
```


> Q18. Using ggplot generate a histogram of this data.

```{r}
ggplot(vax.36) +
  aes(vax.36$percent_of_population_fully_vaccinated ) +
  geom_histogram() +
  xlim(0,1.00)
```

```{r}
vax %>% filter(as_of_date == "2023-03-07") %>%  
  filter(zip_code_tabulation_area=="92040") %>%
  select(percent_of_population_fully_vaccinated)
```


> Q19. Is the 92109 and 92040 ZIP code areas above or below the average value you calculated for all these above?

```{r}
filter(vax.36, zip_code_tabulation_area %in% c("92109", "92040"))
```


```{r}
vax.36.all <- filter(vax, age5_plus_population > 36144)

ggplot(vax.36.all) +
  aes(x = as_of_date, y = percent_of_population_fully_vaccinated, group = zip_code_tabulation_area) +
  geom_line(alpha = 0.2, color = "blue") +
  ylim(0.00,1.00) +
  labs(x="Date", y = "Percent Vaccinated", title="Vaccination Rates Across California", subtitle ="Only areas with populations above 36k are shown") +
  geom_hline(yintercept = avg.36, linetype =2 )

```




