---
title: "Class10_candy"
author: Ashlyn Murphy
format: pdf
---

Candyyyyy data(!!!) with PCA methods

#import the data

```{r}
candy <- read.csv("candy-data.csv", row.names = 1)
candy
```

> Q1. How many different candy types are in this dataset?

```{r}
nrow(candy)
```

> Q2. How many fruity candy types are in the dataset?

```{r}
sum(candy$fruity)
```

> Q3. What is your favorite candy in the dataset and what is it’s winpercent value?

```{r}
candy["Milk Dud", ]$winpercent
```


```{r}
#install.packages("skimr")
```

```{r}
library("skimr")
```

```{r}
skim(candy)
```

> Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

yes the winpercent is on a much larger scale


> Q7. What do you think a zero and one represent for the candy$chocolate column?

whether or not it has chocolate in it

```{r}
library(ggplot2)
```

```{r}
library(dplyr)

```

>Q8. Plot a histogram of winpercent values

```{r}
hist(candy$winpercent)
```

> Q9. Is the distribution of winpercent values symmetrical?

No

> Q10. Is the center of the distribution above or below 50%?
 
below 50%

> Q11. On average is chocolate candy higher or lower ranked than fruit candy?

```{r}
win.choc <- candy[as.logical(candy$chocolate), "winpercent"]

win.fruit <- candy[as.logical(candy$fruity), "winpercent"]

mean(win.choc)

mean(win.fruit)

```

chocolate is higher

> Q12. Is this difference statistically significant?

```{r}
t.test(win.choc, win.fruit)

```

yes

> Q13. What are the five least liked candy types in this set?

```{r}
candy %>% arrange(winpercent) %>% head(5)

```

   
> Q14. What are the top 5 all time favorite candy types out of this set?


```{r}
candy %>% arrange(winpercent) %>% head(5)
```

> Q15. Plot a barplot of winpercent values

```{r}
ggplot(candy) + 
  aes(winpercent, rownames(candy)) +
  geom_col()
```
reordered for trend-spotting:

```{r}
ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col()
```
```{r}
my_cols=rep("black", nrow(candy))
my_cols[as.logical(candy$chocolate)] = "chocolate"
my_cols[as.logical(candy$bar)] = "brown"
my_cols[as.logical(candy$fruity)] = "pink"
```

```{r}
ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col(fill=my_cols) 
```

> Q17. What is the worst ranked chocolate candy?

sixlets

> Q18. What is the best ranked fruity candy?

starburst

#pricepercent

```{r}
#install.packages("ggrepel")
library(ggrepel)

# How about a plot of price vs win
ggplot(candy) +
  aes(winpercent, pricepercent, label=rownames(candy)) +
  geom_point(col=my_cols) + 
  geom_text_repel(col=my_cols, size=3.3, max.overlaps = 5)
```
highest ranked for least money = reese's mini

> What are the top 5 most expensive candy types in the dataset and of these which is the least popular?

Most expensive
```{r}
expensive <- candy %>% arrange(pricepercent) %>% head(5)
expensive
```

least popular

```{r}
least_pop <- candy %>% arrange(pricepercent) %>% head(5)

expensive[which.min(least_pop$winpercent),]



```

```{r}
#install.packages("corrplot")
library(corrplot)
```

```{r}
cij <- cor(candy)
corrplot(cij)
```
> Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?

chocolate and fruity

> Q23. Similarly, what two variables are most positively correlated?

chocolate and winpercent

# PCA

```{r}
pca <- prcomp(candy, scale=T)
summary(pca)
```


PC1 vs PC2 plot

```{r}
plot(pca$x[,1],pca$x[,2], col = my_cols, pch = 16)
```

PCA analysis in ggplot

```{r}
my_data <- cbind(candy, pca$x[,1:3])
```

```{r}
p <- ggplot(my_data) + 
        aes(x=PC1, y=PC2, 
            size=winpercent/100,  
            text=rownames(my_data),
            label=rownames(my_data)) +
        geom_point(col=my_cols)

p
```

```{r}
library(ggrepel)

p + geom_text_repel(size=3.3, col=my_cols, max.overlaps = 7)  + 
  theme(legend.position = "none") +
  labs(title="Halloween Candy PCA Space",
       subtitle="Colored by type: chocolate bar (dark brown), chocolate other (light brown), fruity (red), other (black)",
       caption="Data from 538")
```


```{r}
#install.packages("plotly")
library(plotly)
```

```{r}
#ggplotly(p)
```

```{r}
par(mar=c(8,4,2,2))
barplot(pca$rotation[,1], las=2, ylab="PC1 Contribution")
```

> Q24. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you?

fruity, hard & pluribus which makes sense cause most fruity candy is hard and comes in packages of many candies